---
title: "Linear Mixed Models"
subtitle: "Impact of Age and Education on Clinical Dementia Rating (CDR)"
author: "Bess Tyler (Advisor: Dr. Seals)"
date: '`r Sys.Date()`'
toc: true
toc-depth: 4
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

As the population of older adults continues to increase globally,
cognitive decline has become a significant concern, impacting not only
the individuals affected but also their families and society as a whole.
The factors contributing to cognitive deterioration are multifaceted,
including age, health conditions, lifestyle choices, socioeconomic
status, and more. Understanding the interaction of these influences is
crucial for effectively addressing the issue. Traditional statistical
techniques often struggle to capture these complex relationships, which
limits their ability to provide a comprehensive perspective on cognitive
aging \[4,6\].

An effective approach to this challenge is the use of Linear Mixed
Models (LMMs), which enable a more sophisticated analysis of complex
data structures. These models can account for both fixed effects, which
reflect general trends within populations, and random effects, which
capture individual variability \[2,6\]. This characteristic makes LMMs
particularly valuable for the analysis of longitudinal data, where
multiple observations are collected over time \[3\]. By employing LMMs,
we can more thoroughly investigate how different factors influence
cognitive function \[1, 5\].

The objective of this study is to utilize LMMs to explore the
relationship between cognitive decline and various demographic and
health-related variables, with a specific focus on Clinical Dementia
Rating (CDR) as an indicator of cognitive performance. By analyzing the
interactions between different factors, we aim to develop a clearer
understanding of the elements that contribute to cognitive aging and how
these differ across populations \[7\].

## Methods

::: {.callout-important appearance="minimal"}
Bates stated that " A linear mixed model is described by the
distribution of two vector-valued random variables: Y, the response, and
beta , the vector of random effects. \[2\]
:::

This study employs a quantitative, longitudinal design to investigate
how age and education level influence cognitive decline over time in
older adults, with a primary focus on the effect of age on cognitive
function, measured using the Clinical Dementia Rating (CDR). To analyze
this data, we used Linear Mixed Models (LMMs), which are well-suited for
handling the complexities of repeated measures data and the hierarchical
structure inherent in longitudinal studies \[3, 5\]. LMMs are
particularly advantageous because they allow for the modeling of both
fixed effects, such as the overall influence of age and education level
on cognitive function, and random effects, which account for individual
differences in baseline cognitive scores and trajectories of cognitive
decline \[2, 4\]. Specifically, random intercepts were included in the
model to capture individual differences in starting levels of cognitive
function, while random slopes were used to model the variability in the
rate of cognitive decline over time \[1, 6\].

LMMs are particularly appropriate for this study because cognitive
decline in older adults is often heterogeneous, meaning that individuals
may experience different rates of decline over time. The inclusion of
random effects enables the model to account for this variability, which
is crucial for accurately capturing individual trajectories of cognitive
change \[7\]. Additionally, the use of LMMs helps to mitigate potential
biases from missing data, which is common in longitudinal studies, by
allowing for more flexible handling of incomplete observations \[3, 4\].

### Model Specification

The model structure is as follows:

$Y_{ij} = \beta_0 + \beta_{1ij} + u_{0j} + \epsilon_{ij}$

Where:

-   $\text{Y}_{ij}$ is the response for observation $i$ for subject $j$

-   $\text{Age}_{ij}$ represents the fixed effect of age

-   $u_{0j}$ is the random effect for subject $j$ (accounting for
    individual variability in cognitive decline)

-   $\epsilon_{ij}$ is the residual error term

### Assumptions

-   Linearity - The relationship between the dependent variable and the
    predictors is assumed to be linear.

-   Normality - The residuals are assumed to follow a normal
    distribution.

-   Independence - Residuals should be independent across observations.

-   Homoscedasticity - The variance of the residuals should be constant
    across all levels of the fixed predictors.

## Analysis and Results

### Data Collection

Data was collected from OASIS Longitudinal Dataset, which included a
sample of adults aged 60 - 96. The dataset comprised of 150
participants. Cognitive function was assessed using Clinical Dementia
Rating, while predictors included age, dementia status, and relevant
socioeconomic factors.

### Data Exploration and Visualization

-   Dataset: OASIS-Longitudinal MRI Data in Nondemented and Demented
    Older Adults

    -   Source:
        [NCBI](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2895005/)

    **Summary of the chosen dataset:**

    -   Includes MRI scans of 150 subjects aged 60-96.
    -   Longitudinal data collected over 373 sessions.
    -   Subjects scanned at least twice, with visits separated by at
        least a year.
    -   Clinical Dementia Rating (CDR) categorized as nondemented or
        with mild Alzheimer’s disease.
    -   72 were nondemented throughout, while 64 were demented initially
        and remained so.
    -   Participants were right-handed, consisting of 62 men and 88
        women.
    -   Each session had 3-4 T1-weighted MRI scans.

    Table 1 for this dataset can be seen below:

    ```{r, warning=FALSE, echo=FALSE, message=FALSE}
    #install.packages("tableone")
    #install.packages("bibtex")
    #install.packages("readr")

    #Load libraries
    library(tableone)
    library (rmarkdown)
    library(bibtex)
    library(readr)

    #Load data
    oasis_data <- read_csv("oasis_longitudinal.csv")


    # Recode M/F for easier handling in Table 1 (optional)
    oasis_data$Gender <- ifelse(oasis_data$'M/F' == "M", "Male", "Female")

    # Define the variables to be included in Table 1
    vars <- c("Age", "Gender", "Hand", "EDUC", "SES", "MMSE", "CDR", "eTIV", "nWBV", "ASF")

    # Define the stratifying variable (Group)
    strata <- "Group"

    # Create Table One
    table1 <- CreateTableOne(vars = vars, strata = strata, data = oasis_data, factorVars = c("Gender", "Hand", "CDR"))

    # Print the table with p-values
    print(table1, showAllLevels = TRUE, smd = TRUE)

    ```

    The age distribution of participants is displayed in the visual
    below:

```{r, warning=FALSE, echo=FALSE, message=FALSE}

# Visualization: Age of Participants
counts <- table(oasis_data$Age)
barplot(counts, 
        main = "Age of Participants", 
        xlab = "Age", 
        col = "lightblue", 
        border = "white")
```

The distribution of participants' education levels is presented in the
visual below:

```{r, warning=FALSE, echo=FALSE, message=FALSE}

# Visualization: Age of Participants
counts <- table(oasis_data$EDUC)
barplot(counts, 
        main = "Education Level of Participants", 
        xlab = "Years of Education", 
        col = "lightgreen", 
        border = "white")
```

### Modeling and Results

#### Data Processing and Cleaning

```{r, warning=FALSE, echo=TRUE, message=FALSE}
#install.packages("readr")
library(readr)

oasis_data <- read_csv("oasis_longitudinal.csv")

# Remove duplicate rows
oasis_data <- oasis_data[!duplicated(oasis_data), ]

# Remove rows with missing values
oasis_data <- na.omit(oasis_data)

# View cleaned data
head(oasis_data)

```

The data processing and cleaning steps involved in the provided R code
are as follows:

1.  **Loading the Data**:\
    The first step is loading the necessary library readr, which is used
    to read CSV files efficiently. The read_csv() function is used to
    import the dataset named oasis_longitudinal.csv into R and store it
    in oasis_data.

2.  **Removing Duplicate Rows**:\
    The line oasis_data \<- oasis_data\[!duplicated(oasis_data), \] is
    used to remove duplicate rows from the dataset. The duplicated()
    function checks for duplicate rows, and the ! operator negates it,
    keeping only the unique rows. This ensures that any repeated data
    points are eliminated, preventing redundancy in the analysis.

3.  **Removing Rows with Missing Values**:\
    The na.omit(oasis_data) function is applied to remove any rows that
    contain missing values (NA). This step ensures that the dataset is
    complete for analysis and avoids errors that may arise when handling
    missing data. It eliminates any rows where at least one variable has
    a missing value.

4.  **Viewing the Cleaned Data**:\
    Finally, the head(oasis_data) function is used to display the first
    few rows of the cleaned dataset. This helps the user confirm that
    the data has been processed and cleaned correctly, and it allows for
    a quick visual inspection of the cleaned data.

Overall, these steps ensure that the dataset is free of duplicates and
missing values, making it ready for further analysis.

## Linear Mixed Modeling

### **Model 1**

For the first model, we examined age as the sole predictor of CDR. The
linear mixed model used for model 1 is given by:

$$
\text{CDR}{ij} = \beta_0 + \beta_1 \cdot \text{AGE}{ij} + \mu_{i} + \epsilon_{ij}
$$

```{r, warning=FALSE, echo=T, message=FALSE}
#install.packages("lme4")

library(lme4)
oasis_data$SubjectID <- oasis_data$'Subject ID'
oasis_data$Age <- as.numeric(as.character(oasis_data$Age))

# Fit the linear mixed model with only Age as the predictor
model <- lmer(CDR ~ Age + (1 | SubjectID), data = oasis_data)
summary(model)
```

####  {.smaller}

$$
\text{CDR}{ij} = -.2195 + 0.0066 \cdot \text{AGE}{ij} + \mu_{i} + \epsilon_{ij}
$$

**INTERPRETATION:**

The fixed effect of age shows a positive estimate of 0.0066, suggesting
that as age increases, the CDR score also tends to increase, indicating
greater cognitive decline. The t-value of 2.085 suggests that this
effect of age on cognitive decline is statistically significant. The
intercept of -0.219 and a t-value of -0.889, indicates the baseline CDR
score when age=0 . Although age can not possibly be 0, it provides a
reference point for the model. This equation shows us that age has a
statistically significant positive relationship with CDR, suggesting
that cognitive decline tends to increase slightly with age.

### **Model 2**

For the second model, we expanded our model to include Education Level
(EDUC) as an additional predictor. The linear mixed model used for model
2 is given by:

$$ \text{CDR}{ij} = \beta_0 + \beta_1 \cdot \text{AGE}{ij} + \beta_2 \cdot \text{EDUC}{ij} + \mu_{i} + \epsilon_{ij} $$

```{r, warning=FALSE, echo=T, message=FALSE}
oasis_data$Age <- as.numeric(as.character(oasis_data$Age))
# Fit Linear Mixed Model with Age and Education as the predictors
model <- lmer(CDR ~ Age + EDUC + (1 | SubjectID), data = oasis_data)
summary(model)


```

$$
\text{CDR}{ij} = 0.0403 + 0.0064 \cdot \text{Age}{ij}  -  0.0164 \cdot \text{EDUC}{ij} + \mu_{i} + \epsilon_{ij}
$$

**INTERPRETATION:**

The fixed effect of age is statistically significant, with an estimate
of 0.0064 (t = 2.009), suggesting that as age increases, cognitive
decline, as measured by CDR, also tends to increase. This positive
relationship indicates that older participants generally exhibit higher
CDR scores, reflecting more cognitive decline. In contrast, the fixed
effect of Education Level is not statistically significant, with an
estimate of -0.0164 and a t-value of -1.553. This suggests that
education level, as measured here, does not show a strong or significant
relationship with cognitive decline in this model.

## Conclusion

The results for my analysis reveal age has a statistically significant
effect on Clinical Dementia Rating, with older age associated with
slightly higher dementia severity. Education, while included as a
predictor, does not show a significant effect on CDR in this model. This
suggests that age is a more important factor in cognitive decline, but
Education might not be as influential in this specific dataset.

Linear Mixed Models offer a powerful and flexible approach to analyzing
complex data structures that involve both fixed and random effects. They
allow for more accurate conclusions in the presence of correlations and
varying group-level effects, making them indispensable for research in
many fields \[1, 2, 4, 5\].

## References

\[1\]Agostinelli, C., & Yohai, V. J. (2016). Composite robust estimators
for linear mixed models. *Journal of the American Statistical
Association*, 111(516), 1764-1774.
<https://doi.org/10.1080/01621459.2015.1115358>

\[2\]Bates, D., Mächler, M., Bolker, B., and Walker, S.
(2015), “Fitting Linear Mixed- Effects Models Using lme4,” *Journal of
Statistical Software*, 67,
1–48. <https://doi.org/10.18637/jss.v067.i01>.

\[3\]Booth, J. G., & Hobert, J. P. (1998). Standard Errors of Prediction
in Generalized Linear Mixed Models. *Journal of the American Statistical
Association*, *93*(441), 262–272.
<https://www.tandfonline.com/doi/abs/10.1080/01621459.1998.10474107>

\[4\]Brown, V. A. (2021). An introduction to linear mixed-effects
modeling in R. *Advances in Methods and Practices in Psychological
Science, 4*(1), 1–19. <https://doi.org/10.1177/2515245920960351>

\[5\]McCulloch, C. E. (1996). An introduction to generalized linear
mixed models. In *NSF-CBMS Regional Conference Series in Probability and
Statistics* (Vol. 7). <https://doi.org/10.1214/lnms/1215451825>

\[6\]Sinha, S. K. (2004). Robust Analysis of Generalized Linear Mixed
Models. *Journal of the American Statistical Association*, *99*(466),
451–460. <https://doi.org/10.1198/016214504000000340>

\[7\]Starkweather, J. (2010). Linear mixed effects modeling using R.
*Unpublished Manuscript*. <https://doi.org/10.13140/RG.2.1.3420.4240>
